---
title: "Permutation Testing for ECG Age Gap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permutation Testing for ECG Age Gap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignettes/bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(STA380project)
```

## Overview

This vignette demonstrates the complete workflow implemented in `STA380project`
for comparing ECG-derived age-gap values between atrial fibrillation (AF) and
non-AF groups. The package supports loading a raw CSV, reducing repeated exams
to one exam per patient, computing the age-gap outcome, running a two-sample
permutation test, and visualizing the permutation distribution.

The inferential idea follows the classical randomization logic of permutation
methods discussed by Fisher [@fisher1935]. The vignette itself is organized in
the style recommended for package documentation by Wickham and Bryan
[@wickham2023r].

The age-gap outcome is defined as

\[
\text{age\_gap} = \text{nn\_predicted\_age} - \text{age}.
\]

## Example Data

To keep the vignette self-contained, we create a small CSV file with repeated
exams per patient and then read it with `load_ecg_data()`.

```{r create-example-data}
set.seed(1)

example_df <- data.frame(
  patient_id = c(1, 1, 2, 2, 3, 4, 4, 5),
  exam_id = c(101, 105, 201, 202, 301, 401, 405, 501),
  age = c(61, 61, 54, 54, 70, 48, 48, 66),
  nn_predicted_age = c(64, 67, 51, 55, 77, 46, 50, 69),
  AF = c(TRUE, TRUE, FALSE, FALSE, TRUE, FALSE, FALSE, TRUE),
  extra_column = letters[1:8]
)

tmp_csv <- tempfile(fileext = ".csv")
write.csv(example_df, tmp_csv, row.names = FALSE)

raw_ecg <- load_ecg_data(tmp_csv)
raw_ecg
```

`load_ecg_data()` keeps only the columns needed by the package workflow:
`patient_id`, `exam_id`, `age`, `nn_predicted_age`, and `AF`.

## Step 1: Keep One Exam per Patient

The project proposal correctly noted the repeated-measures issue: multiple ECGs
from the same patient break the independence assumption of a simple two-sample
permutation test. The default workflow therefore keeps one exam per patient.

```{r one-exam-per-patient}
one_exam <- filter_to_unique_patient(raw_ecg, method = "latest")
one_exam
```

In this package, `method = "latest"` keeps the row with the largest `exam_id`
within each patient. That matches the project proposal's recommended default.

## Step 2: Compute the Age-Gap Outcome

```{r compute-age-gap}
one_exam <- compute_age_gap(one_exam)
one_exam[, c("patient_id", "exam_id", "age", "nn_predicted_age", "age_gap", "AF")]
```

A positive `age_gap` means the ECG model predicts an older age than the
patient's actual age; a negative value means the predicted age is younger.

## Step 3: Run the Permutation Test

The package allows three test statistics: mean difference, median difference,
and the Kolmogorov-Smirnov statistic. The mean-difference version is the most
straightforward primary analysis for this vignette.

```{r permutation-test-mean}
x <- one_exam$age_gap[one_exam$AF]
y <- one_exam$age_gap[!one_exam$AF]

perm_mean <- run_permutation_test(x, y, stat = "mean", B = 1000, seed = 123)
perm_mean
```

For completeness, the package also supports a Kolmogorov-Smirnov style
permutation statistic.

```{r permutation-test-ks}
perm_ks <- run_permutation_test(x, y, stat = "ks", B = 500, seed = 123)
perm_ks$t_obs
perm_ks$p_value
```

## Step 4: Plot the Permutation Distribution

```{r permutation-plot}
plot_permutation_distribution(perm_mean$t_perm, perm_mean$t_obs)
```

The histogram shows the null distribution generated by permutation, and the
vertical line marks the observed test statistic.

## Assumptions and Limitations

This workflow is intentionally simple.

1. It assumes that using one exam per patient is an acceptable way to restore
   independence for the two-sample comparison.
2. It does not adjust for covariates such as age, sex, or site.
3. The choice of `exam_id` as a proxy for recency assumes larger values
   correspond to later exams.
4. The KS option is available in the package, but the vignette emphasizes the
   mean-difference statistic because it is easier to interpret substantively.

If the scientific goal requires retaining all repeated exams, then a repeated-
measures model or subject-level resampling scheme would be more appropriate than
this default workflow.

## References

<div id="refs"></div>
